---
title: "HistoricalTorns"
author: "Zoe Schroder"
date: "1/13/2022"
output: html_document
---

Install the packages needed to clean the data. 
```{r}
#install.packages("sf")
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
```

Load the original Significant Tornado data into this session. 
```{r}
SigTor <- st_read("C:/Users/SCHRODEZ/Desktop/Research_Projects/HistoricalTornadoes/SignificantTornadoesData/SignificantTornadoes.shp")
str(SigTor)

#save(SigTor, file = "OriginalTornadoData.RData")

SigTor_geom <- SigTor$geometry #Save the line geometry for later use.
```

Create a new data frame. Convert the time to the appropriate POSIXct format. Create a new column called `LocalTime` in the *SigTor* data. 
%%Correcting time: 
```{r}
#reformat the time column
newdat <- SigTor %>%
          mutate(scaledhour = ifelse( time <= 2300, time + 100, time - 2300),
                 hr_plus1 = format(strptime(substr(as.POSIXct(sprintf("%04.0f", scaledhour), format="%H%M"), 12, 16), '%H:%M'), '%H:%M'),
                hrtime = strptime(hr_plus1, format="%H:%M") - 3600,
                 Time = format(as.POSIXct(hrtime), format = "%H:%M:%S"))

str(newdat)
#create an object with the correctly formatted time
LocalTime <- newdat$Time

#Add the correctly formated time to SigTor
SigTor <- as.data.frame(cbind(SigTor, LocalTime))
```

Get the appropriate convective day for each of these reports. A convective day is a 24-hour period extending from 6 AM to 6 AM. Calculate the casualties, path of the damage area, and abbreviation of the month. 
Separate the date: 
```{r}
SigTor <- SigTor %>%
  mutate(local_DateTime = as.POSIXct(paste(Year, Month, Day, LocalTime), format = "%Y%m%d%H"),
         LocalHour = hour(local_DateTime),
         cDateTime = local_DateTime - as.difftime(6, unit = "hours"),
         cDate = as.Date(as_datetime(ifelse(LocalHour < 6, (local_DateTime - 86400), cDateTime), tz = Sys.timezone())),
         cas = inj + fat,
         AreaPath = len * wid,
         Ma = factor(month.abb[Month], levels = month.abb[1:12])) %>%
  sf::st_sf()
dim(SigTor)[1]

SigTor <- rename(SigTor, time24hr = time)
```
Compute the Energy Dissipation (ED) where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
mag_num <- as.numeric(substr(SigTor$mag,2,2))
ef <- mag_num + 1 
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
SigTor <- SigTor %>%
  mutate(ED = EW3 * AreaPath) #Units of ______
``` 

Get Tornado touchdown (point) data: 
```{r}
SigTor.sfdf <- st_drop_geometry(SigTor)

SigTor.sfdf <- SigTor.sfdf %>%
  st_as_sf(coords = c("slon", "slat"), crs = 4326)
slat <- SigTor$slat
slon = SigTor$slon

SigTor.sfdf <- cbind(SigTor.sfdf, slat, slon)
```

Assign the appropriate timezone to the tornado data using the `tz_lookup` function from the **lutz** package. 
```{r}
#install.packages("lutz")

SigTor.sfdf <- SigTor.sfdf %>% 
  mutate(tz_name = tz_lookup(SigTor.sfdf, method = "accurate"))

timezones <- as.data.frame(tz_list())
US_tz <- as.data.frame(timezones) %>%
  filter(tz_name %in% SigTor.sfdf$tz_name) 
```

Determine whether the DateTime is experiencing standard time or daylight time. Use the `dst` function from the **lubridate** package. 
```{r}
library(lubridate)

SigTor.sfdf <- SigTor.sfdf %>%
  mutate(dst = dst(date))
```

Compute the GMT time for each tornado report using the correct conversion and appropriate timezone. 
```{r}
timezone = c()
Zulu_DateTime = c()

heyya <- for (i in 1:dim(SigTor.sfdf)[1]) {
  tz = (SigTor.sfdf$tz_name[i]) #tz_name column in #28 
    bonjour <- US_tz %>%
      filter(tz_name == tz) 
    bonjour <-  as.data.frame(bonjour)
    
    TZAbbr <- if(SigTor.sfdf$dst[i] == TRUE) {
       bonjour[2,2]
      } else if (SigTor.sfdf$dst[i] == FALSE){
        bonjour[1,2] 
        }  else { 
            NA
        }
    timezone <- append(timezone, TZAbbr)
    
    ZuluTime <- if (SigTor.sfdf$dst[i] == TRUE){
      (SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[2,4])) 
      } else if (SigTor.sfdf$dst[i] == FALSE) {
      (SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[1,4]))
      } else {
      NA
    }
     ZuluTime2 <- strftime(ZuluTime, '%Y-%m-%d %H:%M:%S')
     Zulu_DateTime <- append(Zulu_DateTime, ZuluTime2)
}

SigTor.sfdf <- cbind(SigTor.sfdf, as.data.frame(timezone), as.data.frame(Zulu_DateTime))
```


Convert Zulu_DateTime to a POSIXct object. 
```{r}
SigTor.sfdf$Zulu_DateTime <- as.POSIXct(SigTor.sfdf$Zulu_DateTime, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
```

Standardize all times to CST to remain consistent with the current SPC dataset. 
```{r}
SigTor.sfdf <- SigTor.sfdf %>%
  mutate(Time_CST = Zulu_DateTime - 3600 * 6)
```

Calculate the NARRtime, NARRday, and NARRZtime for each report. 
```{r}
SigTor.sfdf$Zulu_DateTime <- force_tz(SigTor.sfdf$Zulu_DateTime, tzone = "UTC")
NARRtime <- align.time(SigTor.sfdf$Zulu_DateTime, n = (60 * 60 * 3)) - 3600 * 3 ###Had to remove the -3600*3
NARRday = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%Y/%m/%d")
NARRZtime = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%H")

SigTor.sfdf <- cbind(SigTor.sfdf, NARRday, NARRZtime)
```

Establish the geometry for a tornado tracks dataset using the original geometry from the Historical Tornadoes data. 
```{r}
#Tornado Tracks
SigTorTracks.sfdf <- st_as_sf(cbind(st_drop_geometry(SigTor.sfdf), SigTor_geom))
```

Rearrange the data. Create a tornado touchdown (point) dataset. 
```{r}
geom_points <- SigTor.sfdf$geometry

SigTor.sfdf = st_drop_geometry(SigTor.sfdf)


Ma <- SigTor.sfdf$Ma
Time_CST <- SigTor.sfdf$Time_CST
AreaPath <- SigTor.sfdf$AreaPath
ED <- SigTor.sfdf$ED
cas <-  SigTor.sfdf$cas
cDate <- SigTor.sfdf$cDate
Zulu_DateTime <-  SigTor.sfdf$Zulu_DateTime
county <- SigTor.sfdf$county

SigTorns <- cbind(SigTor.sfdf[,1:3],  SigTor.sfdf[,11:13], Ma, SigTor.sfdf[,17:19], Time_CST, county, SigTor.sfdf[,5:9], AreaPath, ED, cas, SigTor.sfdf[,28:29], SigTor.sfdf[,14:15], cDate, Zulu_DateTime, SigTor.sfdf[,30:32], SigTorTracks.sfdf[,35:36], geom_points )
```

Create a tornado tracks dataset. 
```{r}
geom_tracks <- SigTorTracks.sfdf$geometry
SigTorTracks.sfdf = st_drop_geometry(SigTorTracks.sfdf)

Ma <- SigTorTracks.sfdf$Ma
Time_CST <- SigTorTracks.sfdf$Time_CST
AreaPath <- SigTorTracks.sfdf$AreaPath
ED <- SigTorTracks.sfdf$ED
cas <-  SigTorTracks.sfdf$cas
cDate <- SigTorTracks.sfdf$cDate
Zulu_DateTime <-  SigTorTracks.sfdf$Zulu_DateTime
county <- SigTorTracks.sfdf$county

SigTornTracks <- cbind(SigTorTracks.sfdf[,1:3],  SigTorTracks.sfdf[,11:13], Ma, SigTorTracks.sfdf[,17:19], Time_CST, county, SigTorTracks.sfdf[,5:9], AreaPath, ED, cas, SigTorTracks.sfdf[,28:29], SigTorTracks.sfdf[,14:15], cDate, Zulu_DateTime, SigTorTracks.sfdf[,30:32], SigTorTracks.sfdf[,35:36], geom_tracks)
```

```{r}
#save(SigTorns, SigTornTracks, file = "HistoricalTornadoes.RData")
```


