---
title: "HistoricalTorns"
author: "Zoe Schroder"
date: "1/13/2022"
output: html_document
---

Install the packages needed to clean the data. 
```{r}
#install.packages("sf")
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
```

Load the original Significant Tornado data into this session. 
```{r}
SigTor <- st_read("E:/Zoe/Projects/HistoricalTornadoes/SignificantTornadoesData/1880-1990-torn-aspath.shp")
```

```{r}
str(SigTor)

#save(SigTor, file = "OriginalTornadoData.RData")

SigTor_geom <- SigTor$geometry #Save the line geometry for later use.
```

Get the appropriate convective day for each of these reports. A convective day is a 24-hour period extending from 6 AM to 6 AM. Calculate the casualties, path of the damage area, and abbreviation of the month. 
Separate the date: 
```{r}
SigTor <- SigTor %>%
  mutate(DateTime = as.POSIXct(paste(date, time), format="%Y-%m-%d %H:%M:%S"), 
         cas = inj + fat,
         AreaPath = len * wid,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
dim(SigTor)[1]
```
Compute the Energy Dissipation (ED) where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
mag_num <- as.numeric(substr(SigTor$mag,2,2))
ef <- mag_num + 1 
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
SigTor <- SigTor %>%
  mutate(ED = EW3 * AreaPath) #Units of ______
``` 

Get Tornado touchdown (point) data: 
```{r}
SigTor.sfdf <- st_drop_geometry(SigTor)

SigTor.sfdf <- SigTor.sfdf %>%
  st_as_sf(coords = c("slon", "slat"), crs = 4326)
slat <- SigTor$slat
slon = SigTor$slon

SigTor.sfdf <- cbind(SigTor.sfdf, slat, slon)
```

Calculate the Convective day!:
```{r}
SigTor.sfdf <- SigTor.sfdf %>%
  mutate(cDateTime = DateTime - as.difftime(6, unit = "hours"),
         hour = hour(DateTime),
         cDate = as.Date(as_datetime(ifelse(hour < 6, (DateTime - 86400), cDateTime), tz = Sys.timezone()))) %>%
  sf::st_sf()
```



Calculate the NARRtime, NARRday, and NARRZtime for each report. 
```{r}
SigTor.sfdf$Zulu_DateTime <- force_tz(SigTor.sfdf$DateTime, tzone = "UTC")
NARRtime <- align.time(SigTor.sfdf$Zulu_DateTime, n = (60 * 60 * 3)) + 3600 *3 ###Had to remove the -3600*3
NARRday = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%Y/%m/%d")
NARRZtime = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%H")

SigTor.sfdf <- cbind(SigTor.sfdf, NARRday, NARRZtime)
```

Establish the geometry for a tornado tracks dataset using the original geometry from the Historical Tornadoes data. 
```{r}
#Tornado Tracks
SigTornTracks<- st_as_sf(cbind(st_drop_geometry(SigTor.sfdf), SigTor_geom))
SigTorns <- SigTor.sfdf
```

```{r}
#save(SigTorns, SigTornTracks, file = "HistoricalTornadoes.RData")
```


