---
title: "HistoricalTorns"
author: "Zoe Schroder"
date: "1/13/2022"
output: html_document
---

Install the packages needed to clean the data. 
```{r}
#install.packages("sf")
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
```
Load the original Significant Tornado data into this session. 
```{r}
SigTor <- st_read("C:/Users/SCHRODEZ/Desktop/Research_Projects/HistoricalTornadoes/SignificantTornadoesData/SignificantTornadoes.shp")
str(SigTor)
```
Correcting time: 
```{r}
newdat <- SigTor %>%
          mutate(scaledhour = ifelse( time <= 2300, time + 100, time - 2300),
                 hr_plus1 = format(strptime(substr(as.POSIXct(sprintf("%04.0f", scaledhour), format="%H%M"), 12, 16), '%H:%M'), '%H:%M'),
                hrtime = strptime(hr_plus1, format="%H:%M") - 3600,
                 Time = format(as.POSIXct(hrtime), format = "%H:%M:%S"))

str(newdat)
Time <- newdat$Time

SigTor <- as.data.frame(cbind(SigTor, Time))
```
Separate the date: 

```{r}

SigTor <- SigTor %>%
  mutate(DateTime = as.POSIXct(paste(Year, Month, Day, Time), format = "%Y%m%d%H"),
         Hour = hour(DateTime),
         cDateTime = DateTime - as.difftime(6, unit = "hours"),
         cDate = as.Date(as_datetime(ifelse(Hour < 6, (DateTime - 86400), cDateTime), tz = Sys.timezone())),
         cas = inj + fat,
         AreaPath = len * wid,
         Ma = factor(month.abb[Month], levels = month.abb[1:12])) %>%
  sf::st_sf()
dim(SigTor)[1]

SigTor <- rename(SigTor, time24hr = time)
```

where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
mag_num <- as.numeric(substr(SigTor$mag,2,2))
ef <- mag_num + 1 
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
SigTor <- SigTor %>%
  mutate(ED = EW3 * AreaPath) #Units of ______
``` 

```{r}
SigTor.sfdf <- st_drop_geometry(SigTor)

SigTor.sfdf <- SigTor.sfdf %>%
  st_as_sf(coords = c("slon", "slat"), crs = 4326)
SigTor.sfdf <- cbind(SigTor.sfdf, SigTor$slat, SigTor$slon)


install.packages("lutz")
library(lutz)
timezones <- as.data.frame(tz_list())

SigTor.sfdf <- SigTor.sfdf %>% 
  mutate(tz_name = tz_lookup(SigTor.sfdf))

test <- as.data.frame(timezones) %>%
  filter(tz_name %in% SigTor.sfdf$tz_name) 

Mo_DT <- c(4:10)
Mo_ST <- c(1:2, 12)
Day_DST <- c(13, 6)

TZ_Abbr = c()

heyya <- for (i in 1:dim(SigTor.sfdf)[1]) {
  tz = st_drop_geometry(SigTor.sfdf[i, 28]) 
    bonjour <- test %>%
      filter(tz_name == tz$tz_name) 
    bonjour <-  as.data.frame(bonjour)
    
    TZAbbr <- if (SigTor.sfdf$Month[i] %in% Mo_DT) {
      bonjour[2,2]
    } else if (SigTor.sfdf$Month[i] %in% Mo_ST) {
      bonjour[2,1]
    } else if (SigTor.sfdf$Month[i] == 3 & SigTor.sfdf$Day[i] > 13){
      bonjour[2,2]
    } else if (SigTor.sfdf$Month[i] == 3 & SigTor.sfdf$Day[i] <= 13){
      bonjour[2,1]
    }  else if (SigTor.sfdf$Month[i] == 11 & SigTor.sfdf$Day[i] > 6){
      bonjour[2,2]
    } else if (SigTor.sfdf$Month[i] == 11 & SigTor.sfdf$Day[i] <= 6){
      bonjour[2,1]
    }
    TZ_Abbr <- append(TZ_Abbr, TZAbbr)

  }
  
  
```



tz_list()










```{r}
EST <- c("CT", "DE", "DC", "GA", "ME", "MD", "MA", "NH", "NJ", "NY", "NC", "OH", "PA", "RI", "SC", "VT", "VA", "WV")
CST <- c("AL", 'AR', "IL", "IA", "LA", "MN", "MS", "MO", "OK", "WI")
AKST <- c("AK")
MST <- c("AZ","CO","MT", "NM", "UT", "WY")
PST <- c("CA", "WA")


```

```{r}
FL_CST <- c("ESCAMBIA", "SANTA ROSA", "ESCAMBIA /SANTA ROSA", "ESCAMBIA /SANTA ROSA. FL /ESCAMBIA. AL", "ESCAMBIA ,FL /ESCAMBIA. AL", "OKALOOSA", "WALTON", "HOLMES", "WASHINGTON", "BAY", "JACKSON", "CALHOUN", "BAY /CALHOUN /JACKSON", "GULF", "HOLMES /WASHINGTON", "JACKSON. FL /HOUSTON. AL", "SANTA ROSA /OKALOOSA")
FLtorns <- SigTor %>%
  filter(st == "FL") %>%
  mutate(tz = ifelse(county %in% FL_CST, "CST", "EST"))
```

```{r}
ID_PST <- c("BENEWAH", "BONNER", "BOUNDARY", "CLEARWATER", "KOOTNAI", "LATAH", "LEWIS", "NEZ PERCE", "SHOSHONE", "IDAHO")
IDtorns <- SigTor %>%
  filter(st == "ID") %>%
  mutate(tz = ifelse(county %in% ID_PST, "PST", "MST"))
```

```{r}
IN_CST <- c("LAKE", "PORTER", 'LAPORTE', 'STARKE', 'NEWTON', "JASPER", "KNOX", "DAVIESS", "MARTIN", "PIKE", "DUBOIS", "GIBSON", "POSEY", "VANDERBURGH", "WARRICK", "SPENCER", "PERRY", "GIBSON /PIKE", "	KNOX /DAVIESS", "	LAKE /NEWTON", "LAKE /PORTER", 	"LAKE /PORTER /LAPORTE", "VANDERBURGH /GIBSON", "VANDERBURGH/WARRICK")
INtorns <- SigTor %>%
  filter(st == "IN") %>%
  mutate(tz = ifelse(county %in% IN_CST, "CST", "EST"))
```

```{r}
KS_MST <- c("GREELEY", "HAMILTON", "SHERMAN", "WALLACE", "SHERMAN /THOMAS")
KStorns <- SigTor %>%
  filter(st == "KS") %>%
  mutate(tz = ifelse(county %in% KS_MST, "MST", "CST"))
```

```{r}
KY_CST <- c("ADAIR", "ALLEN", "BALLARD", "BARREN", "BRECKINRIDGE", "BUTLER", "CALDWELL", "CALLOWAY", "CARLISLE", "CHRISTIAN", "CLINTON", "CRITTENDEN", "CUMBERLAND", "DAVIESS", "EDMONSON", "FULTON", "GRAVES", "GRAYSON", "GREEN", "HANCOCK", "HART", "HENDERSON", "HICKMAN", "HOPKINS", "LIVINGSTON", "LOGAN", "LYON", "MCCRACKEN", "MCLEAN", "MARSHALL", "METCALFE", "MONROE", "MUHLENBERG", 'OHIO', "RUSSELL", "SIMPSON", "TODD", "TRIGG", "UNION", "WARREN", "WEBSTER", "ALIEN /BARREN", "BARRON /METCALFE", "BUTLER /WARREN", "CALDWELL /HOPKINS", "CALLOWAY /MARSHALL", "CALLOWAY /MARSHALL /TRIGG /CALDWELL /CHRISTIAN", "CHRISTIAN /TODD", "CLINTON /WAYNE", "CRITTENDEN /WEBSTER /HENDERSON", "CUMBERLAND /CLINTON /WAYNE", "DAVIES /HANCOCK", "DAVIES. KY /SPENCER. IN", "DAVIESS /HANCOCK, KY /PERRY. IN	
", GRAVES /MARSHALL, GRAYSON /HARDIN, 	GREEN /ADAIR, GREEN /TAYLOR, HENDERSON /DAVIESS, 	HICKMAN /GRAVES, HOPKINS /MUHLENBERG /OHIO, LIVINGSTON /LYON /CALDWELL, MARSHALL /LYON /CALDWELL, MARSHALL /LYONS, MCCRACKEN. KY    /MASSAC /POPE. IL /LIVINGSTON CRITTENDEN /WEBSTER /MCLEAN /DAVIESS, KY, 	MCCRACKEN. KY /MASSAC. IL, MONROE /CUMBERLAND /ADAIR /RUSSELL, 	METCALF /ADAIR, MUHLENBERG /BUTLER, MUHLENBERG /BUTLER /WARREN, MUHLENBERG /OHIO, 	OHIO /GRAYSON /BRECKINRIDGE /HARDIN )
KYtorns <- SigTor %>%
  filter(st == "KY") %>%
  mutate(tz = ifelse(county %in% KY_CST, "CST", "EST"))
```

```{r}
MI_CST <- c("DICKINSON", "GOGEBIC", "IRON", "MENOMEE")
```

```{r}
NE_MST <- c("ARTHUR", "BANNER", "BOX BUTTE", "CHASE", "CHERRY", "CHEYENNE", "DAWES", "DEUEL", "DUNDY", "GARDEN", "GRANT", "HOOKER", "KEITH", "KIMBALL", "MORRILL", "PERKINS", "SCOTTS BLUFF", "SHERIDAN", "SIOUX")
```

```{r}
NV_MST <- c("ELKO")
```

```{r}
ND_MST <- c("BOWMAN", "ADAMS", "SLOPE", "HETTINGER", "GRANT", "STARK", "BILLINGS", "GOLDEN VALLEY")
```

```{r}
OR_MST <- c("MALHEUR")
```

```{r}
SD_MST <- c("BENNETT", "BUTTE", "CORON", "CUSTER", "DEWEY", "FALL RIVER", "HAAKON", "HARDING", "JACKSON", "HONES", "LAWRENCE", "MEADE", "MELLETTE", "PENNINGTON", "PERKINS", "SHANNON", 'STANLEY', "TODD", "ZIEBACH")
```

```{r}
TN_EST <- c("ANDERON", "BLOUNT", "BRADLEY", 'CAMPBELL', "CARTER", "CLAIBORNE", "COCKE", 'GRAINGER', "GREENE", "HAMBLEN", "HAMILTON", "HANCOCK", "HAWKINS", "JEFFERSON", "JOHNSON", "KNOX", "LOUDON", "MCMINN", "MEIGS", "MONROE", "MORGAN", "POLK", "RHEA", "ROANE", "SCOTT", "SEVIER", "SULLIVAN", "UNICOI", "UNION", "WASHINGTON")
```

```{r}
TX_MST <- c("EL PASO", "HUDSPETH")
```






