---
title: "HistoricalTorns"
author: "Zoe Schroder"
date: "1/13/2022"
output: html_document
---

Install the packages needed to clean the data. 
```{r}
#install.packages("sf")
suppressMessages(library(sf))
suppressMessages(library(dplyr))
suppressMessages(library(lubridate))
suppressMessages(library(lutz))
suppressMessages(library(xts))
suppressMessages(library(chron))
```

Load the original Significant Tornado data into this session. 
```{r, eval = FALSE}
SigTor <- st_read("C:/Users/SCHRODEZ/Desktop/Research_Projects/HistoricalTornadoes/SignificantTornadoesData/SignificantTornadoes.shp")
str(SigTor)

SigTor2 <- SigTor
```
Correcting time: 
```{r, eval = FALSE}
newdat <- SigTor %>%
          mutate(scaledhour = ifelse( time <= 2300, time + 100, time - 2300),
                 hr_plus1 = format(strptime(substr(as.POSIXct(sprintf("%04.0f", scaledhour), format="%H%M"), 12, 16), '%H:%M'), '%H:%M'),
                hrtime = strptime(hr_plus1, format="%H:%M") - 3600,
                 Time = format(as.POSIXct(hrtime), format = "%H:%M:%S"))

str(newdat)
LocalTime <- newdat$Time

SigTor <- as.data.frame(cbind(SigTor, LocalTime))
```
Separate the date: 

```{r, eval = FALSE}

SigTor <- SigTor %>%
  mutate(local_DateTime = as.POSIXct(paste(Year, Month, Day, LocalTime), format = "%Y%m%d%H"),
         LocalHour = hour(local_DateTime),
         cDateTime = local_DateTime - as.difftime(6, unit = "hours"),
         cDate = as.Date(as_datetime(ifelse(LocalHour < 6, (local_DateTime - 86400), cDateTime), tz = Sys.timezone())),
         cas = inj + fat,
         AreaPath = len * wid,
         Ma = factor(month.abb[Month], levels = month.abb[1:12])) %>%
  sf::st_sf()
dim(SigTor)[1]

SigTor <- rename(SigTor, time24hr = time)
```

where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r, eval = FALSE}
perc <- c(1, 0, 0, 0, 0, 0, 
          .772, .228, 0, 0, 0, 0,
          .616, .268, .115, 0, 0, 0,
          .529, .271, .133, .067, 0, 0,
          .543, .238, .131, .056, .032, 0,
          .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
mag_num <- as.numeric(substr(SigTor$mag,2,2))
ef <- mag_num + 1 
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
SigTor <- SigTor %>%
  mutate(ED = EW3 * AreaPath) #Units of ______
``` 

```{r, eval = FALSE}
SigTor.sfdf <- st_drop_geometry(SigTor)

SigTor.sfdf <- SigTor.sfdf %>%
  st_as_sf(coords = c("slon", "slat"), crs = 4326)
slat <- SigTor$slat
slon = SigTor$slon

SigTor.sfdf <- cbind(SigTor.sfdf, slat, slon)
```

```{r, eval = FALSE}
#install.packages("lutz")

SigTor.sfdf <- SigTor.sfdf %>% 
  mutate(tz_name = tz_lookup(SigTor.sfdf))

timezones <- as.data.frame(tz_list())
US_tz <- as.data.frame(timezones) %>%
  filter(tz_name %in% SigTor.sfdf$tz_name) 
```

```{r, eval = FALSE}
Mo_DT <- c(4:10)
Mo_ST <- c(1:2, 12)
Day_DST <- c(13, 6)

timezone = c()
Zulu_DateTime = c()

heyya <- for (i in 1:dim(SigTor.sfdf)[1]) {
  tz = st_drop_geometry(SigTor.sfdf[i,28]) #tz_name column in #28 
    bonjour <- US_tz %>%
      filter(tz_name == tz$tz_name) 
    bonjour <-  as.data.frame(bonjour)
    
    TZAbbr <- if (SigTor.sfdf$Month[i] %in% Mo_DT) {
      bonjour[2,2]
    } else if (SigTor.sfdf$Month[i] %in% Mo_ST) {
      bonjour[1,2]
    } else if (SigTor.sfdf$Month[i] == 3 & SigTor.sfdf$Day[i] > 13){
      bonjour[2,2]
    } else if (SigTor.sfdf$Month[i] == 3 & SigTor.sfdf$Day[i] <= 13){
      bonjour[1,2]
    }  else if (SigTor.sfdf$Month[i] == 11 & SigTor.sfdf$Day[i] > 6){
      bonjour[2,2]
    } else if (SigTor.sfdf$Month[i] == 11 & SigTor.sfdf$Day[i] <= 6){
      bonjour[1,2]
    }
    timezone <- append(timezone, TZAbbr)

    ZuluTime <- if (SigTor.sfdf$Month[i] %in% Mo_DT) {
      SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[2,4])
    } else if (SigTor.sfdf$Month[i] %in% Mo_ST) {
      SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[1,4])
    } else if (SigTor.sfdf$Month[i] == 3 & SigTor.sfdf$Day[i] > 13){
      SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[2,4])
    } else if (SigTor.sfdf$Month[i] == 3 & SigTor.sfdf$Day[i] <= 13){
      SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[1,4])
    }  else if (SigTor.sfdf$Month[i] == 11 & SigTor.sfdf$Day[i] > 6){
      SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[2,4])
    } else if (SigTor.sfdf$Month[i] == 11 & SigTor.sfdf$Day[i] <= 6){
      SigTor.sfdf$local_DateTime[i] - (3600 * bonjour[1,4])
    }
    Zulu_DateTime <- append(Zulu_DateTime, ZuluTime)    
}

SigTor.sfdf <- cbind(SigTor.sfdf, as.data.frame(timezone), as.data.frame(Zulu_DateTime))
```

```{r, eval = FALSE}
SigTor.sfdf <- SigTor.sfdf %>%
  mutate(Time_CST = Zulu_DateTime - 3600 * 6)
```


```{r, eval = FALSE}
SigTor.sfdf$Zulu_DateTime <- force_tz(SigTor.sfdf$Zulu_DateTime, tzone = "UTC")
NARRtime <- align.time(SigTor.sfdf$Zulu_DateTime, n = (60 * 60 * 3)) - 3600 * 3 ###Had to remove the -3600*3
NARRday = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%Y/%m/%d")
NARRZtime = format(as.POSIXct(strptime(NARRtime,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%H")

SigTor.sfdf <- cbind(SigTor.sfdf, NARRday, NARRZtime)
```

```{r, eval = FALSE}
#Tornado Start Points
#save(SigTor.sfdf, file = "HistoricalTornadoTouchDowns.RData")

#Tornado Tracks
SigTorTracks.sfdf <- cbind(SigTor2, SigTor.sfdf[19:32])
#save(SigTor.sfdf, file = "HistoricalTornadoTracks.RData")
```

```{r}
load("HistoricalTornadoTouchDowns.RData")
load("HistoricalTornadoTracks.RData")
```



